\bf Employee Attrition: What Makes an Employee Quit?

### 2018-01-09 LatexPaper/paper.tex
Added final paper

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subfig}
\graphicspath{ {Images/} }
\usepackage{float}
\usepackage[backend=bibtex,style=verbose-trad2]{biblatex}
%\bibliography{bib}

@@

\title{Employee Attrition: What makes an employee leave?}
\author{
	Christopher Boomhower$^1$
	\and
	Stacey Fabricant$^2$
	\and
	Alex Frye$^1$
	\and
	David Mumford$^2$
	\and
	Michael Smith$^1$
	\and
	Lindsay Vitovsky$^1,^2$
}
\date{}
\maketitle{}

@@

%definition of quote with addition of [\textbf{Abstract --- ]
\renewenvironment{abstract}
               {\list{}{\rightmargin\leftmargin}%
                \item[\textbf{\hspace{10mm}Abstract ---}]\relax}
               {\endlist}

\begin{abstract}\noindent In this paper, we present a model for employee attrition prediction and discuss the ethical impacts of using such a model within private and public sector organizations. As it is in Human Resource personnel’s best interest to improve retention, implementing statistical and machine learning techniques is the most viable means to attrition abatement. To this end, we examine Office of Personnel Management public sector, Bureau of Labor Statistics public sector, and IBM anonymized private sector employee separation data. Three classification models (Methodologies include Logistic Regression, Random Forest, and K Nearest Neighbor) are trained and tested on these data before selecting our best fit model for attrition prediction. We finally use metrics such as Gini and Permutation Importance to identify the most impactful variables in determining prediction outcome before presenting the ethical ramifications of using such outputs in HR planning. [WILL ADD SENTENCE FOR MAIN RESULT AND SENTENCE FOR MAIN CONCLUSION ONCE THESE DEVELOPMENTS ARE COMPLETE].\end{abstract}

@@

\paragraph{How much does it really cost to lose an employee? Studies such as the Center’s for American Progress analysis (November, 2012) indicate a separated employee may cost anywhere between 16 percent to 213 percent depending on the position [1]. Precisely quantifying this may seem out of reach depending on the complexity of a particular role, but areas of impact that one may foresee at many organizations are: 1) determining if the employee’s vacancy should be replaced or duties handed off to others; 2) posting the job opportunity to various outlets; 3) interviewing, hiring, and training a replacement; 4) enduring lowered employee morale / possible lower productivity from remaining employees; and 5) tolerating a lower skill set from an underdeveloped replacement [2].}

\subparagraph{Corporations are keenly aware of the downsides to losing employees and exert great effort to maintain retention levels. In their efforts to not only attract talented workers but retain them as well, businesses provide substantial benefits [3]. With industry competitiveness the norm, many employers may still face retention challenges as their employees have alternate employment options. To become even more proactive in attrition prevention, companies must gain a solid understanding for the reasons their employees separate. Foresight into attrition development and contributing factors empowers Human Resource departments to improve retention efforts through improved planning and intervention. While such insights are available to organizations that store employee data, these understandings are not within reach without sufficient analysis.}

\subparagraph{The first step in gaining foresight into employee attrition is obtaining pertinent data. Companies are understandably reluctant to release the methods, proprietary or purchased, that use even anonymous data to help them in their management of human resources. Various articles allude to this challenge [4, 5, 6]. However, we identified three valid sources of Human Resources data in the forms of Office of Personnel Management data, Bureau of Labor Statistics data, and the “IBM HR Analytics Employee Attrition” dataset. All three forms were analyzed in unison to complement one another in insight and model validity.}

\subparagraph{During analysis, dimensionality reduction was performed on the datasets. This was essential to reduce numerous correlating and covariant relationships present between dataset variables. Only after these relationships were addressed and the datasets simplified were the data prepared for modeling.}

\subparagraph{[ADD PARAGRAPH ABOUT MODELLING PHASE AS PROJECT DEVELOPMENT PROGRESSES][ADD PARAGRAPH SUMMARIZING MAIN RESULTS AS PROJECT DEVELOPMENT PROGRESSES][ADD PARAGRAPH SUMMARIZING MAIN CONCLUSIONS AS PROJECT DEVELOPMENT PROGRESSES][ADD PAPER OVERVIEW PARAGRAPH ONCE REMAINING PAPER CONTENT FILLS IN]}

@@

\paragraph{The U.S. Office of Personnel Management (OPM) serves as the central Human Resources department for all Federal agencies, including the management of federal agency health insurance and retirement benefits. Their oversight of policy implementation as well as being a general resource for all agency Human Resource departments makes their employment data of particular interest for this paper. OPM releases, on a quarterly basis, updated separation data on its over two million federal civil service workers.  We used these anonymized data sets to create a full year of separation data, spanning October 2014 - September 2015.   These data [7], include the following variables: 1) age group, captured categorically in 5 year increments (AGELVL and AGELVLT), 2) agency type and subtype information (AGYTYP, AGYTYPT, AGY, AGYT, AGYSUB, and AGYSUBT), 3) gender (GENDER and GENDERT), 4) General Schedule Grade, or equivalent, dictating salary level (GSEGRD, 5) geographical location of the employee down to the state level (LOCTYP, LOCTYPT, LOC, LOCT), 5)length of service as a government employee captured in one year increments until five years was achieved, then every five after that (LOSLVL and LOSLVLT), 6)occupation type  in regards to white or blue collar, occupation family, and in some cases, down to the particular occupation itself (i.e. "chaplain") (OCCTYP, OCCTYPT, OCCFAM, OCCFAMT, OCC, OCCT), 7) occupation category among seven main categories (i.e. "Technical" and "Clerical")(PATCO and PATCOT), 8) pay plan and grade for both GSE and non-GSE occupations (PPTYP, PPTYPT, PPGROUP, PPGROUPT, PAYPLAN, PAYPLANT, and PPGRD), 9) salary range, captured categorically starting with $<$ \$20,000 and followed in \$10,000 increments (SALLVL and SALLVLT), 10) reason for separation (SEP and SEPT), and 11) type of appointment and worschedules, denoting factors such as permanence, seasonality, and executive status (TOATYP, TOATYPT, TOA, TOAT, WSTYP, WSTYPT, WORKSCH, and WORKSCT).   }

\subparagraph{There are certain points worth noting to understand our findings in the correct context.  The General Schedule Pay Scale is the centralized pay scale used by many civil service agencies. Whether or not a person is paid under the GS, the OPM converts the level he or she is paid under to the GS for data collection purposes. The scale numbers 1-15 and there are ten “steps” within each of those levels.  Also, to remain competitive with industry salaries, the GS operates a locality adjustment scale that adds a particular percentage to a person’s salary, based on city of occupation alone.}

\subparagraph{ Also, beginning on January 1, 1987,  new civil service employees were paid under the Federal Employees Retirement System (FERS). This mix of Social Security, a Basic Benefit Plan, and a Thrift Savings Plan helps make civil service positions stand out from many current private sector jobs in terms of retention. We account for this factor in our research, taking into consideration the confounding effect this has on length of service (LOS). At the time of this paper, the pension is based on salary and length of service.}

@@

\paragraph{Our research focuses on the following employee types, only using observations that fit the following rubric: 1) domestic workers (no international positions), 2) employees aged 20 and greater, 3)employees with a job grade level greater than 7 (i.e. white collar only), and only those who are considered full-time employees (i.e. no contracted, temporary, or part-time employees).}

@@

\subparagraph{After reviewing the percentage representation of the various occupation types, as well as investigating the types of jobs that comprise each grouping, we kept only Professional and Administrative positions. Fig. 1 depicts the overall record counts present within our dataset by occupation category type. As observed, there are more employees of Administrative and Professional occupation types than any other category. Reviewing separation type ratios which comprise these total counts provides further insight and justification into why Professional and Administrative jobs are targeted within this paper.}

\begin{figure}
\caption{Fig. 1. Occupation Category Type – Observation Counts}
\end{figure}

\subparagraph{The percentage plot of Fig. 2[Field] indicates the distribution of separation types within each occupation category. Note for improved granularity, non-separated employee data has been removed from the visualization as there are far more employees remaining at work than there are leaving, which therefore skews plotted results. So, only true separations are represented visually.}

\begin{figure}
\caption{Fig. 2. Occupation Category Type – Separation Percentages with NS Removed}
\end{figure}

\subparagraph{As Fig. 2[Field] illustrates, approximately the same number of Professional employees who retired (SD) in this timeframe quit (SC) as well. Treating retirement as a proxy indicator for retention, this even distribution of quits vs. retirements suggests great retention rates among Professional employees and aligns with non-separation counts as well. Similar argument is to be made of Blue Collar jobs also, and the combination of individual transfers out (SA) and quits is approximately the same in size as retirements for Administrative types. However, the Professional category type was identified as a better candidate for primary model application due to its large total count, as discussed previously, in combination with its more widely diverse professional job types that more closely resemble private sector occupations. Nonetheless, due to its high employee count, strong retention numbers, and similar transfer and quit counts compared to retirements, the Administrative occupation category was selected as secondary target for modeling purposes as will be discussed further in section 5. Therefore, data for all occupation category types, except Professional and Administrative, were removed from our dataset.}

@@

\paragraph{Six new attributes were created through aggregation or calculation: 1) SEP Count by Date \& Occupation – Total number of separations (of any type) for a given Date and Occupation; 2) SEP Count by Date \& Location – Total number of separations (of any type) for a given Date and Location; 3) Industry Average Salary – Average salary amongst non-separated employees, grouped by quarter, occupation, pay grade, and work schedule; 4) Lower Limit Age – Youngest age within each age level category; 5) Years to Retirement – Based on FERS retirement eligibility baseline of 57 years of age [8]; and 6) Salary Over/Under Industry Average – Difference between computed average salary of non-separated employees and actual salary for each observation. Note also another 1,293 observations were removed after calculating industry average salary as they had no matching non-separation observations (matched on quarter, occupation type, pay plan /grade, and work schedule), which were utilized to ensure realistic salary averages.}

@@

\paragraph{In addition to the OPM data, we merged 10 attributes from the Bureau of Labor Statistics (BLS). Data were sourced from Federal Government industry codes across all regions. Although assumed to be highly correlated, we sourced both Level (Total number) and Rate (Percentage of Level to total employment and / or job openings) for the following statistics: 1) Job Openings, 2) Layoffs, 3) Quits, 4) Total Separations, and 5) Other Separations. While Rate paints an aggregated, holistic picture for job market trends, Level provides a raw count for total separations alone. Both these statistics were captured by a monthly aggregate and merged to the OPM data by their respective months.}

@@

\paragraph{From an original 8,423,336 observations, we ultimately result in two datasets for Professional and Administrative occupations, reduced to 1,282,291 and 2,128,093 observations respectively. Of the 1,282,291 observations present in the reduced OPM dataset of professional occupations, only 23,008 contained actual separation data and all other observations were considered non-separation. This state of data was inoperable for analysis; therefore, a sample design was determined to mitigate sample size constraints and high variance amongst frequencies of separation types.}

\subparagraph{Data were divided into groups based on separation type, allowing a maximum of 4,000 observations per type to persist forward during analysis. In so doing, the following retirement separation types were combined: 1) SD Retirement – Voluntary, 2) SE Retirement – Early Out, 3) SF Retirement – Disability, and 4) SG Retirement – Other. Next, the following separation types were dropped completely: 1) SB Transfer Out – Mass Transfer, 2) SK Death, 3) SL Other Separation, and 4) SJ Termination (Expired Appt/Other).}

\subparagraph{Within each separation group (including non-separation), proportional allocation was performed on a combination of date and age level strata to ensure a sample demographic which, as closely as possible, represents that of the original strata-level populations. After sampling, we were left with 16,638 observations in our primary dataset for model generation. The secondary dataset of Administrative occupations also contained these unbalanced strata, and thus the same sample design was implemented to reduce the dataset further to 17,238 observations.}

@@

\begin{figure}
\caption{Fig. 3. Numeric Attribute and Separation Type Correlation Matrix}
\end{figure}

\subparagraph{As indicated in Fig. 3, there are several intuitive observations to be made of the correlation values shared by separation type and the remaining quantitative attributes present in the dataset. Note numerical values shown are Pearson’s correlation coefficient values. Separation due to voluntary retirement (SEP\_SD) is positively correlated with paygrade (r = 0.2), salary differential from the industry (r = 0.3), and length of service (r = 0.6). Even though these are to be expected, this is still worth noting as it supports the idea that the prospect of retiring with benefits, as opposed to quitting, can help entice an employee to stay. Of course, the pension calculation is heavily based on number of years of service, so it is unsurprising that length of service is the highest positively correlated factor for separation by voluntary retirement.}

\subparagraph{The correlation values for separation from quitting (SEP\_SC) in Fig. 3 is what interests us most.  As paygrade decreases, the likelihood of separation by quitting increases (r = 0.2). This makes logical sense, as paygrade is directly tied to income.  However, this correlation also exactly matches the correlation for the salary differential from the industry, so there is not enough direction to say that the status of the paygrade is what matters more so than the actual salary. Though not indicated in Fig. 3, salary and paygrade level do in fact strongly correlate with one another (r = 0.9). As such, the two are most likely one and the same in an employee's mind, and only one of the two will be used for modeling purposes.}

\subparagraph{Finally, the correlation of quits to years until retirement eligibility is positively correlated (r = 0.3), whereas it is negatively correlated with length of service (r = -0.4); the further someone is from being able to retire, the more likely they are to quit. But when considering the correlations observed for paygrade and salary differential, a key takeaway is that the job positions for which it is more difficult to replace and train should be assigned higher paygrades and lower income differentials from industry. This conclusion aside, these insights solidified our data as being predictive of attrition.}

\subparagraph{[WILL DESCRIBE CATEGORICAL DATA RELATIONSHIPS WITH SEPARATION]}

@@

\paragraph{Methodologies Logistic Regression, Random Forest, and K Nearest Neighbor(KNN) were implemented to predict separation types amongst our sampled professional occupations. Four disparate feature input selections were utilized in attempts to reduce dimensionality in our initial input of 99 attributes. Ultimately, Logistic Regression produced most effective trained results and the generated models were applied to a different demographic of Administrative occupations.}

@@

\paragraph{Four feature input selections were made during the modeling process in search of the best features for each model type individually. Selections made were: 1) Full 99 Raw Scaled Value Features; 2) Principal Component Analysis(PCA) Eigenvector Matrix; 3) Top 15 Raw Scaled Features within Each PCA Loading; 4) Manual Logistic Regression Feature Selection.}

\subparagraph{Full 99 Raw Scaled Value Features: All 99 values after Min Max Scaling was performed.}

\subparagraph{Principal Component Analysis(PCA) Eigenvector Matrix: Principal component analysis was performed on our professional occupation samples in attempts to reduce our 99 input attributes. As can be seen in Figure ~\ref{fig:PCA_Scree_CumVar}, the percentage of explained variance leveled out around 22 components, at which point cumulative explained variance approached 80\%.}

@@

% Comment(Chris): Missing Top 15 Raw Scaled Features within Each PCA Loading here
\subparagraph{These 22 principal components were utilized both as a dataset of component vectors as well as to identify the top 15 features within each of the 22 component loading magnitudes. The latter approach reduced our inputs to 49 features, allowing only the most important features to proceed into our model as raw scaled values, based on PCA component loadings.}

\subparagraph{Manual Logistic Regression Feature Selection: Important independent variables were identified by first assessing multicollinearity and covariance among the 99 raw features, and then iteratively generating logistic regression models - removing one feature at a time with each iteration. Removal criteria were based on Variance Inflation Factor (VIF) values (Threshold for removal was VIF of 10), Null Deviance, Log Likelihood, Akaike Information Criterion (AIC), Bayesian Information Criterion (BIC), and p-values ($\alpha = 5$). Only statistically significant features remained for model input after completing this process (32 variables in total).}

@@

\paragraph{[DISCUSS CV SPLIT FOLDS HERE.....]}

\subparagraph{Among the three model types and four separate input features tested, Logistic Regression with Manual Feature Selection was identified as most effective. As depicted in Table ~\ref{fig:AccTable} the [INTERPRET MODEL MEAN ACCURACY RESULTS HERE]}

@@

\subparagraph{Furthermore, we may see visually in Figure ~\ref{fig:AccPlot}, the variation in accuracy for each model across the 5 fold tests further supports the Logistic Regression results as a strong selection.[INTERPRET FOLD ACCURACY METRICS HERE, MAKING NOTE THAT LEGEND INDEXES CORRESPOND TO Table ~\ref{fig:AccTable}]}

\subparagraph{[discuss lr roc plot here.....Figure ~\ref{fig:ROCPlot}.]}

\subparagraph{}
\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{ROCPlot.jpg}
\caption{Accuracies Across Test Folds for Top 4 Model Results}
\label{fig:ROCPlot}
\end{figure}

@@

\paragraph{[CHRIS TO PLACE TEXT HERE.]}

\subparagraph{}
\begin{figure}
\centering
\subfloat[first]{
  \includegraphics[width=60mm]{Violin_AGELVL_H.pdf}
}
\subfloat[second]{
  \includegraphics[width=60mm]{Violin_AGELVL_G.pdf}
}
\hspace{0mm}
\subfloat[third]{
  \includegraphics[width=60mm]{Violin_AGELVL_I.pdf}
}
\subfloat[fourth]{
  \includegraphics[width=60mm]{Violin_AGELVL_B.pdf}
}
\hspace{0mm}
\subfloat[fifth]{
  \includegraphics[width=60mm]{Violin_AGELVL_F.pdf}
}
\caption{Violin Plots - Significant Age Levels}
\label{fig:ViolinAge}
\end{figure}

>>>>

\documentclass[10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage [english]{babel}
\usepackage [autostyle, english = american]{csquotes}
\MakeOuterQuote{"}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage[all]{nowidow}
\graphicspath{ {Images/} }
\usepackage{float}
\usepackage[belowskip=-20pt,aboveskip=0pt,font=normalsize]{caption}
\usepackage[backend=bibtex]{biblatex}
\bibliography{bib}

\DeclareFieldFormat{labelnumberwidth}{\mkbibbold{#1\adddot}}

@@

\title{\bf Employee Attrition: What Makes an Employee Quit?}
\author{
	Christopher Boomhower$^1$
	\and
	Stacey Fabricant$^2$
	\and
	Alex Frye$^1$
	\and
	David Mumford$^2$
	\and
	Michael Smith$^1$
	\and
	Lindsay Vitovsky\textsuperscript{1,2}
}
\date{}
\maketitle{}

@@

%definition of quote with addition of [\textbf{Abstract --- ]
\renewenvironment{abstract}
               {\list{}{\rightmargin\leftmargin}%
                \item[\textbf{\hspace{10mm}Abstract.}]\relax}
               {\endlist}

\begin{abstract}\fontsize{9}\noindent Albeit to varying degrees, employee attrition is a costly challenge faced by many employers \cite{kantor}.  In this paper, we present a model for predicting employee attrition, as well as discuss the serious ethical implications of using such a model within organizations. To accomplish this, we examined publicly available data from the Office of Personnel Management, the Bureau of Labor Statistics, and IBM.  With these sources, we determined a set of statistically significant factors that correlate to an employee’s decision to quit, and determined to which types of occupations our model may be applied.  After applying Principal Component Analysis and classification methods K-Nearest Neighbors and Random Forest, it was Logistic Regression that allowed us to simplify the model and predict employee quits with the highest accuracy of our testing methods, achieving a greater than seventy-four percent success rate.  \end{abstract}

@@

\paragraph{How much does it really cost to lose an employee? Studies such as the Center’s for American Progress analysis (November, 2012) indicate a separated employee may cost between 16 to just over 213 percent of their annual salary, depending on the position \cite{boushey}. Precisely quantifying this may seem out of reach depending on the complexity of a particular role, but easily foreseeable areas of impact are: 1) determining if an employee’s vacancy should be replaced or duties handed off to others; 2) posting the job opportunity to various outlets; 3) interviewing, hiring, and training a replacement; 4) enduring lowered employee morale and possible lower productivity from 
 remaining employees; and 5) tolerating a lower skill set from an underdeveloped replacement \cite{kantor}.}

\subparagraph{Of course, many corporations are keenly aware of the downsides to losing employees.  As such, many exert great efforts to maintain retention levels, such as by providing substantial workplace benefits \cite{glassdoor}. Gaining an understanding of the reasons why employees separate certainly empowers Human Resource departments to improve retention through improved planning and intervention.  While such insights are available to organizations that store employee data, these understandings are not within reach without sufficient analysis.}

\subparagraph{The first step in gaining foresight into employee attrition is obtaining pertinent data. Companies are understandably reluctant to release the methods, proprietary or purchased, that use even anonymous data to help them in their management of human resources. Various articles allude to this challenge \cite{abellin}, \cite{richter}, \cite{walker}. However, we identified three valid sources of Human Resources data in the forms of Office of Personnel Management data, Bureau of Labor Statistics data, and the "IBM HR Analytics Employee Attrition" data set. All three forms were analyzed in unison to complement one another in insight and model validity.}

\subparagraph{Our process began with data sampling and traditional exploratory data analysis, but we quickly determined we would need to narrow our focus on these very large data sets which include many different types of variables.  We limited our scope to U.S. domestic, professional (white collar) jobs that have a General Schedule Equivalent of Level 7 or above (this is discussed in further detail within Section 2).  We also investigated the qualitative nature of these data to ensure that, at each step of data removal or dimensionality reduction, we were remaining on course to answer our ultimate question of interest, "What makes an employee quit?" Effects beyond the control of the employer or the employee (e.g. death) were carefully reviewed to properly apply and interpret our modeling techniques.}

\subparagraph{Our initial modeling techniques were classification methods that incorporated all variables, as we were still investigating which variables affected the decision to quit with statistical significance.  After eventually simplifying the data by keeping variables with high correlation to attrition responses and only focusing on responses within control of the employer (i.e. employers cannot control disabilities, deaths, etc), the models began to greatly improve.  Ultimately, it was logistic regression that provided us the most success, both intuitively and statistically.}

@@

\paragraph{The U.S. Office of Personnel Management (OPM) serves as the central Human Resources department for all federal agencies, including the management of federal agency health insurance and retirement benefits. Their oversight of policy implementation, as well as being a general resource for all agency Human Resource departments, makes their employment data of particular interest for this paper. OPM releases, on a quarterly basis, updated separation data on its over two million federal civil service workers.  We used these anonymized data sets to create a full year of separation data, spanning October 2014 - September 2015.   These data \cite{opm}, include the following variables: 1) age group, captured categorically in 5-year increments (AGELVL and AGELVLT); 2) agency type and sub-type information (AGYTYP, AGYTYPT, AGY, AGYT, AGYSUB, and AGYSUBT); 3) gender (GENDER and GENDERT); 4) General Schedule Grade, or equivalent, dictating salary level (GSEGRD); 5) geographical location of the employee down to the state level (LOCTYP, LOCTYPT, LOC, LOCT); 6) length of service as a government employee captured in one-year increments until five years is achieved, then every five after that (LOSLVL and LOSLVLT); 7) occupation type  in regards to white or blue collar, occupation family, and in some cases, drilled down to the particular occupation itself (e.g. "chaplain") (OCCTYP, OCCTYPT, OCCFAM, OCCFAMT, OCC, OCCT); 8) occupation category among seven main categories (e.g. "Technical" and "Clerical") (PATCO and PATCOT); 9) pay plan and grade for both General Schedule Equivalent (GSE) and non-GSE occupations (PPTYP, PPTYPT, PPGROUP, PPGROUPT, PAYPLAN, PAYPLANT, and PPGRD); 10) salary range, captured categorically starting with less than \$20,000 and increasing in \$10,000 increments (SALLVL and SALLVLT); 11) reason for separation (SEP and SEPT); and 12) type of appointment and work schedules, denoting factors such as permanence, seasonality, and executive status (TOATYP, TOATYPT, TOA, TOAT, WSTYP, WSTYPT, WORKSCH, and WORKSCT).}

\subparagraph{There are certain points worth noting to understand our findings in the correct context.  The General Schedule (GS) Pay Scale is the centralized pay scale used by many civil service agencies. Even if a person is not paid under the GS, the OPM converts the level he or she is paid under to the GS for data collection purposes. The scale numbers 1-15, and there are ten "steps" within each of these levels.  Also, to remain competitive with industry salaries, the GS operates a locality adjustment scale that adds a particular percentage to a person’s salary, based on city of occupation alone.}

\subparagraph{Also, beginning on January 1, 1987,  new civil service employees were paid under the Federal Employees Retirement System (FERS). This mix of Social Security, a Basic Benefit Plan, and a Thrift Savings Plan helps make civil service positions stand out from many current private sector jobs in terms of retention. We account for this factor in our research, taking into consideration the confounding effect this has on length of service (LOS). At the time of this paper, the pension is based on salary and length of service.}

@@

\paragraph{Our research focuses on the following employee types, only using observations that fit the following rubric: 1) domestic workers (no international positions), 2) employees aged 20 and greater, 3) employees with a job grade level greater than 7 (i.e. white collar only), and 4) only those who are considered full-time employees (i.e. no contracted, temporary, or part-time employees).}

@@

\paragraph{After reviewing the percentage representation of the various occupation types, as well as investigating the types of jobs that comprise each grouping, we kept only Professional and Administrative positions. Figure~\ref{fig:PATCOTOccCounts} depicts the overall record counts present within our data set by occupation category type. As observed, there are more employees of Administrative and Professional occupation types than any other category. Reviewing separation type ratios which comprise these total counts provides further insight and justification into why Professional and Administrative jobs are targeted within this paper.}

\subparagraph{}
\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{PATCOTOccCounts.png}
\caption{Occupation Category Type – Observation Counts}
\label{fig:PATCOTOccCounts}
\end{figure}

\subparagraph{The percentage plot of Figure~\ref{fig:PATCOTSepPercentages} indicates the distribution of separation types within each occupation category. Note for improved granularity, non-separated employee data has been removed from the visualization as there are far more employees remaining at work than there are leaving, which therefore skews plotted results. So, only true separations are represented visually.}

\subparagraph{}
\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{PATCOTSepPercentages.png}
\caption{Occupation Category Type – Separation Percentages with NS Removed}
\label{fig:PATCOTSepPercentages}
\end{figure}

\subparagraph{As Figure \ref{fig:PATCOTSepPercentages} illustrates, approximately the same number of Professional employees who retired (SD) in this time frame quit (SC) as well. Treating retirement as a proxy indicator for retention, this even distribution of quits vs. retirements suggests great retention rates among Professional employees and aligns with non-separation counts as well. Similar argument is to be made of Blue Collar jobs also, and the combination of individual transfers out (SA) and quits is approximately the same in size as retirements for Administrative types. However, the Professional category type was identified as a better candidate for primary model application due to its large total count, as discussed previously, in combination with its more widely diverse professional job types that more closely resemble private sector occupations. Nonetheless, due to its high employee count, strong retention numbers, and similar transfer and quit counts compared to retirements, the Administrative occupation category was selected as secondary target for modeling purposes as will be discussed further in Section 5. Therefore, data for all occupation category types, except Professional and Administrative, were removed from our data set.}

@@

\paragraph{Seven new attributes were created through aggregation or calculation: 1) SEP Count by Date \& Occupation – Total number of separations (of any type) for a given Date and Occupation; 2) SEP Count by Date \& Location – Total number of separations (of any type) for a given Date and Location; 3) Industry Average Salary – Average salary among non-separated employees, grouped by quarter, occupation, pay grade, and work schedule; 4) Lower Limit Age – Youngest age within each age level category; 5) Years to Retirement – Based on FERS retirement eligibility baseline of 57 years of age \cite{bls}; 6) Length of Service Square Root Transformation – Square root transformation to account for service length's skewed distribution; and 7) Salary Over/Under Industry Average – Difference between computed average salary of non-separated employees and actual salary for each observation. Another 1,293 observations were removed after calculating industry average salary, as they had no matching non-separation observations (matched on quarter, occupation type, pay plan/grade, and work schedule), which were utilized to ensure realistic salary averages.}

@@

\paragraph{In addition to the OPM data, we merged 10 attributes from the Bureau of Labor Statistics (BLS). Data were sourced from Federal Government industry codes across all regions. Although assumed to be highly correlated, we sourced both Level (Total number) and Rate (Percentage of Level to total employment and/or job openings) for the following statistics: 1) Job Openings, 2) Layoffs, 3) Quits, 4) Total Separations, and 5) Other Separations. While Rate paints an aggregated, holistic picture for job market trends, Level provides a raw count for total separations alone. Both these statistics were captured by a monthly aggregate and merged to the OPM data by their respective months.}

@@

\paragraph{Data were divided into groups based on separation type, allowing a maximum of 7,500 observations per type to persist forward during analysis. In so doing, the following retirement separation types were combined to create our Non-Separation (NS) group: 1) NS, Non-Separation; 2) SA, Transfer Out - Individual Transfer; 3) SB, Transfer Out - Mass Transfer; 4) SD, Retirement - Voluntary; 5) SE, Retirement - Early Out; 6) SF, Retirement - Disability; and 7) SG, Retirement - Other. Our Quit (SC) group is stand-alone in our analysis to serve as our main focus and goal for prediction. All other separation types were removed for analysis as they were either irrelevant to our goals or were due to uncontrollable events, such as death.}

\subparagraph{Within each separation group (including non-separation), proportional allocation was performed on a combination of date and age level strata to ensure a sample demographic which, as closely as possible, represents that of the original strata-level populations. After sampling, we were left with 14,920 observations in our primary data set for model generation. The secondary data set of Administrative occupations also contained these unbalanced strata, and thus the same sample design was implemented to reduce the data set to 14,918 observations.}

@@

\subparagraph{}
\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{SEP_CorrMatrix.jpg}
\caption{Numeric Attribute and Separation Type Correlation Matrix}
\label{fig:SEP_CorrMatrix}
\end{figure}

\subparagraph{As indicated in Figure~\ref{fig:SEP_CorrMatrix}, there are several intuitive observations to be made of the correlation values shared by separation type and the remaining quantitative attributes present in the data set. Note numerical values shown are Pearson’s correlation coefficient values. The most noteworthy mention identified in the correlation matrix is that Separation (SEP\_NS) is positively correlated with length of service (r = 0.6). Though expected, this is still worth noting as it supports the prospect of retiring with benefits, as opposed to quitting, can help entice an employee to stay. Of course, the pension calculation is heavily based on number of years of service, so it is unsurprising that length of service is the highest positively correlated factor for non-separated employees.}

\subparagraph{The correlation values for separation from quitting (SEP\_SC) in Figure~\ref{fig:SEP_CorrMatrix} is what interests us most.  As general schedule grade (GSEGRD) decreases, the likelihood of separation by quitting increases (r = -0.3). This makes logical sense, as schedule grades are directly tied to income.  However, this correlation also exactly matches the correlation for the employee salary, so there is not enough direction to say that the status of the schedule grade matters moreso than the actual salary. Though not indicated in Figure~\ref{fig:SEP_CorrMatrix}, salary and schedule level do, in fact, strongly correlate with one another (r = 0.9). As such, the two are most likely one-and-the-same in an employee's mind, and only one of the two will be used for modeling purposes.}

\subparagraph{Finally, the correlation of quits to years until retirement eligibility is positively correlated (r = 0.4); the further someone is from being able to retire, the more likely they are to quit. But when considering the correlations observed for schedule grade and salary, a key takeaway is that the job positions for which it is more difficult to replace and train should be assigned higher schedule grades and lower income differentials from industry. This conclusion aside, these insights solidified our data as being predictive of attrition.}

@@

\paragraph{Methodologies Logistic Regression, Random Forest, and K-Nearest Neighbor (KNN) were implemented to predict separation types amongst our sampled professional occupations. Four disparate feature input selections were utilized in attempts to reduce dimensionality in our initial input of 99 attributes. Ultimately, Logistic Regression produced most effective trained results and the generated models were applied to a different demographic of Administrative occupations.}

@@

\paragraph{Four feature input selections were made during the modeling process in search of the best features for each model type. Selections made were: 1) Full 99 Raw Scaled Value Features; 2) Principal Component Analysis (PCA) Eigenvector Matrix; 3) Top 15 Raw, Scaled Features within each PCA loading; and 4) Manual Logistic Regression Feature Selection.  These four selections are further explained below:}

\subparagraph{\textit{Full 99 Raw Scaled Value Features}: All 99 values after Min-Max Scaling were included.}

\subparagraph{\textit{Principal Component Analysis Eigenvector Matrix}: PCA was performed on our professional occupation samples in attempts to reduce our 99 input attributes. As can be seen in Figure~\ref{fig:PCA_Scree_CumVar}, the percentage of explained variance leveled out around 22 components, at which point cumulative explained variance approached 80\%. These 22 principal components were utilized as a data set of component vectors.}

@@

\subparagraph{\textit{Top 15 Raw Scaled Features within Each PCA Loading}: Utilizing the PCA analysis performed above, the top 15 features within each of the 22 components were identified through absolute loading magnitudes. This reduced our inputs to 46 features, allowing only the most important features to proceed into our model as raw, scaled values based on PCA loadings.}

\subparagraph{\textit{Manual Logistic Regression Feature Selection}: Important independent variables were identified by first assessing multicollinearity and covariance among the 99 raw features, and then iteratively generating logistic regression models - removing one feature at a time with each iteration. Removal criteria were based on Variance Inflation Factor (VIF) values (Threshold for removal was VIF of 10), Null Deviance, Log Likelihood, Akaike Information Criterion (AIC), Bayesian Information Criterion (BIC), and p-values ($\alpha = 0.05$). Only statistically significant features remained for model input after completing this process (32 variables in total).}

@@

\paragraph{To train our model, we utilized Stratified K-Fold Cross Validation for our classification analysis, with five folds. From our original sample size of 14,922, each fold, or iteration, was split approximately 20\% as test observations, utilizing the rest as training observations all while keeping the ratio of classes equal among separation types. These folds allowed us to validate that we are not overfitting our model and increase our understanding of whether our model may hold up for external data sets.}

\subparagraph{Among the three model types and four separate input features tested, Logistic Regression with manual feature selection was identified as having the greatest overall accuracy. As depicted in Table~\ref{fig:AccTable}, the winning model produced an average accuracy of 74.595\%, closely followed by Random Forest using the same feature inputs at 74.119\%.}

@@

\subparagraph{Furthermore, we see visually in Figure~\ref{fig:AccPlot} that the variation in accuracy for each model across the five fold tests further supports the Logistic Regression results as a strong selection. Legend indexes, matching those found in Table~\ref{fig:AccTable}, show Logistic regression as the winner among 2/5 iterations, and a close second only trailing behind 0.4020\% and 0.4692\% in iteration 2 and 3, respectively. Interestingly, the disparate feature inputs to the Random Forest model created vastly different results, with the top 15 PC features producing very inconsistent results across folds. With large differences in results between the first two folds, we push further into True Positive / False Positive rates to ensure we do not have too large of inconsistencies in each fold test.}

\subparagraph{To assess accuracies of the Logistic Regression model further, we produced a Receiver Operating Characteristic (ROC) Curve plot as may be seen in Figure~\ref{fig:ROCPlot}. This plot displays the True Positive / False Positive rates across test predictions for each fold. Although Fold 0 appears to be a very low area curve, it is not overly concerning given there is no unexpected variation in the curve and it aligns with our accuracy results reviewed previously. Of our final tests, we push forward with our analysis using the Logistic Regression model with manually selected features.}

\subparagraph{}
\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{ROCPlot.jpg}
\caption{Logistic Regression Model Train ROC Curve across 5 folds}
\label{fig:ROCPlot}
\end{figure}

@@

\paragraph{While we have briefly discussed the methods used to identify the most impactful variables to be included in our Logistic Regression model, we now discuss these variables and their real-world significance for Human Resource organizations. Only by understanding these features' influence on attrition may organizations hope to curb the rate of voluntary employee separation.}
